Observation of a Magnetic-Torque Enhanced by Interacting Spin Ensembles

### 2021-02-22 Mes papiers/Méca CR/18112020/SI/Meca_cross_SI.tex

\author{C. Pellet-Mary, P. Huillery, M. Perdriat, G. H\'etet}

@@

\section{NV$^-$ center Theory}
\subsection{NV spin hamiltonian}
The hamiltonian of the electronic spin of the negatively charged nitrogen-vacancy center in its ground state can be written as :
\begin{equation*}
  \hat{\mathcal{H}}_s=\hbar D S_{z'}^2 + \hbar \gamma_e \textbf{B}\cdot\hat{\textbf{S}},
  \end{equation*} 
where $D = (2\pi) 2.87$ GHz is the crystal field splitting originating from spin-spin interactions, and $\gamma_e = 28 $GHz/T is the electron gyromagnetic ratio. 
The \textbf{z'} axis in the $S_{z'}$ operator here is the axis formed by the nitrogen atom and the vacancy in the body fixed frame.
We neglect contributions from the strain and local electric field in the hamiltonian since we are working with magnetic fields on the order of 0.1T, which induce splittings  larger than the splitting of the zero-field ESR line ($\approx 20$ MHz). We also neglect the hyperfine interaction with the nuclear spin of the $^{14}$N atom since we are working with ensembles with typical inhomogeneous broadening of 5~MHz.

@@

\subsection{Diamond crystalline axes and degeneracy conditions}
\begin{figure}[!ht]
  \centering \scalebox{0.45}{\includegraphics{SI_Cristallo}}
  \caption{a) Representation of the four crystalline axes of the diamond b) Representation of the magnetic field in the diamond crystalline basis c) Representation of the crystalline planes in the $(\theta , \phi)$ basis. The [110]$_\perp$ family of planes are shown using dashed lines, [100]$_\perp$ using plain lines. The [100] direction is marked by red circles, the [111] direction by green squares. The magnetic field path in the experiment of Fig. 3 of the main text is shown with blue dots.}
	\label{cristallo}
\end{figure}
There are four possible crystalline axes for the N-V direction (so-called ``classes" of NV) in the diamond. They are depicted in Fig. \ref{cristallo} b) and correspond to the crystalline directions [$111$], [$1\bar 1 \bar 1$], [$\bar 1 1 \bar 1$] and [$\bar 1 \bar 1 1$].

The magnetic field direction is represented in Fig. \ref{cristallo} a), where the usual polar and azimuthal angles $\theta$ and $\phi$ are defined with respect to the \textbf{z'} ([$001$]) direction. For some orientations of the magnetic field, the projection of the magnetic field on two or more NV axes will be identical, and therefore the energy level of the corresponding classes will be the same. These degeneracies are represented in Fig \ref{cristallo} c), where the dashed lines are the locii of the crystalline planes orthogonal to the [110] directions (6 planes in total). When the magnetic field belongs to these planes, we observe a degeneracy between two classes of NVs, as can be seen in the Fig. 4 of the main paper.

The plain lines are the locii of the planes orthogonal to the [100] directions (3 planes in total). When the magnetic field lies in these planes, all classes are co-resonant, as can be seen in Fig. \ref{CR_22}.
The red circles correspond to the [100] directions, for which the four classes of NV are degenerate. The green squares correspond to the [111] direction where one class is aligned with the magnetic field, and the three others are degenerate. Finally the blue dots correspond to the path followed by the magnetic field in the experiment presented in Fig 3 of the main text, where we can see that a plane from the [110]$_\perp$ family is being crossed.

@@

Another way to estimate the torque is depicted in Fig \ref{Torque1classe}. Under green excitation and in the presence of an external magnetic field, the spins will acquire a magnetization $\gamma_e \langle\hat{\mathbf S}\rangle$ which, under the low magnetic fields ($<$ 200 G) we are working at, will be oriented at an angle of 90$^\circ$ from the NV axis : $\langle \hat S_{z'} \rangle \approx 0$ and $\langle \hat S_\perp \rangle \neq 0$. This magnetization vanishes when the magnetic field is aligned with the NV center since there is no longer a transverse field responsible for the mixing of the eigenstates.

@@

The amplitude of the torque with respect to the magnetic field orientation (at a B field amplitude of 100 G) is represented in 1D in Fig \ref{Torque1classe}-c) where we can see a behavior very close to $|\sin(2\theta)|$, which is the formula (3) of the main text that we obtained through a perturbative approach. The same torque amplitude is represented in 2D in Fig \ref{Torque1classe} d). The two purple dots in the map correspond to the [111] direction when the magnetic field is aligned with the centers. The curvy line corresponds to the $[111]_\perp$ plane. Importantly, the maximum torque value is $1\cdot 10^{-26}$ N.m for a single spin.

@@

Fig 3 a) in the main text, represents the same map as Fig \ref{Torque1classe}-c) but including the four NV centers, one in each of the possible [111] orientations. We can see that the maximum torque actually decreased to $3\cdot 10^{-28}$ N.m even though we increased the number of NV centers by four. This is due to the directional averaging of the torque generated by the four centers. The torque per NV center is decreased by more than two orders of magnitude when taking the directional averaging into consideration.
Fig 3 b) in the main text shows the same map, this time taking into account the modification of the spin lifetime due to cross-relaxations. The detailed model is presented in the section \ref{Simu}.
There are two things to note here : 
\begin{enumerate}
\item The maximum torque has increased by an order of magnitude compared to the previous case. It reached up to $3\cdot 10^{-27}$ N.m for four spins, so about $10^{-27}$ Nm per spin. Qualitatively, this is because cross-relaxation will lower the torque contribution of specific classes (the ones that get depolarized), meaning that the end result is closer to the single spin case (there is less directional averaging).
\item The change in magnetic torque is resonant, it happens only when different classes are brought to resonance. This can be seen by comparing Fig 3 b) of the main text to the [110]$_\perp$ planes that were drawn in Fig. \ref{cristallo}. The change in the signal when scanning a magnetic field across a CR will be much sharper than the sinusoidal change in the spin-torque.
\end{enumerate}

@@

It should be noted that the main uncertainty comes here from the diamond size can change the expected result by an order of magnitude.

>>>>

\author{C. Pellet-Mary$^1$, P. Huillery$^1$, M. Perdriat$^1$, G. H\'etet$^1$}

@@

\section{NV$^-$ center Theory}
\subsection{NV spin hamiltonian}
The hamiltonian of the electronic spin of the negatively charged nitrogen-vacancy center in its ground state can be written as :
\begin{equation*}
  \hat{\mathcal{H}}_s=\hbar D S_{z'}^2 + \hbar \gamma_e \textbf{B}\cdot\hat{\textbf{S}},
  \end{equation*} 
where $D = (2\pi) 2.87$ GHz is the crystal field splitting originating from spin-spin interactions, and $\gamma_e = 28 $GHz/T is the electron gyromagnetic ratio. 
The \textbf{z'} axis in the $S_{z'}$ operator here is the axis formed by the nitrogen atom and the vacancy in the body fixed frame.
We neglect contributions from the strain and local electric field in the hamiltonian since we are working with magnetic fields on the order of 10 mT, which induce splittings larger than the splitting of the zero-field ESR line ($\approx 20$ MHz). We also neglect the hyperfine interaction with the nuclear spin of the $^{14}$N atom since we are working with ensembles with typical inhomogeneous broadening of 5~MHz.

@@

\subsection{Diamond crystalline axes and degeneracy conditions}
\begin{figure}[!ht]
  \centering \scalebox{0.45}{\includegraphics{SI_Cristallo}}
  \caption{a) Representation of the four crystalline axes of the diamond b) Representation of the magnetic field in the diamond crystalline basis c) Representation of the crystalline planes in the $(\theta , \phi)$ basis. The $\{ 110 \}$ family of planes are shown using dashed lines and the $\{ 110 \}$ family using plain lines. The [100] direction is marked by red circles, the [111] direction by green squares. The magnetic field path in the experiment of Fig. 3 of the main text is shown with blue dots.}
	\label{cristallo}
\end{figure}
There are four possible crystalline axes for the N-V direction (so-called ``classes" of NV) in the diamond. They are depicted in Fig. \ref{cristallo} b) and correspond to the crystalline directions [$111$], [$1\bar 1 \bar 1$], [$\bar 1 1 \bar 1$] and [$\bar 1 \bar 1 1$].

The magnetic field direction is represented in Fig. \ref{cristallo} a), where the usual polar and azimuthal angles $\theta$ and $\phi$ are defined with respect to the \textbf{z'} ([$001$]) direction (we denote with ' the axes in the diamond frame). For some orientations of the magnetic field, the projection of the magnetic field on two or more NV axes will be identical, and therefore the energy level of the corresponding classes will be the same. These degeneracies are represented in Fig \ref{cristallo} c), where the dashed lines are the locii of the $\{ 110 \}$ family of planes (plane normal to the [110] direction and all other equivalent directions, making 6 planes in total). When the magnetic field belongs to these planes, we observe a degeneracy between two classes of NVs, as can be seen in the Fig.\ref{scan_dege} or in Fig. 4 of the main paper.

The plain lines are the locii of the $\{ 100 \}$ family of planes (3 planes in total). When the magnetic field lies in these planes, all classes are co-resonant, as can be seen in Fig.\ref{scan_dege} or in Fig.\ref{CR_22}.
The red circles correspond to the [100] directions, for which the four classes of NV are degenerate. The green squares correspond to the [111] direction where one class is aligned with the magnetic field, and the three others are degenerate. Finally the blue dots correspond to the path followed by the magnetic field in the experiment presented in Fig 3 of the main text, where we can see that a plane from the $\{ 110 \}$ family is being crossed.

\section{Depolarization induced by NV-NV cross-relaxation}
As explained in the main text, when the density of NV$^-$ centers in the sample is high enough (typically for concentrations higher than 1 ppm), the ensemble of NV spins will lose some of its polarization through dipolar coupling between the NV centers. This phenomenon is at the heart of the mechanism that allows us to exalt the magnetic suceptibility of our diamond through dipolar interaction, and it has already been observed independently by many groups in bulk diamond \cite{Jarmola} \cite{mrozek_longitudinal_2015} \cite{choi_depolarization_2017} \cite{akhmedzhanov_microwave-free_2017} \cite{giri_coupled_2018}

In particular, \cite{choi_depolarization_2017} proposes a model based on "fluctuators" : a subgroup of NV centers with a very short lifetime (possibly due to their electron tunnelling in and out of the NV site) that can act as a source of classical noise with a central frequency given by the usual transition frequencies of the NV$^-$ spin Hamiltonian. A prediction of this model is that the modified lifetime of the ensemble of NV centers should have a stretch exponential profile (of the form $e^{-\sqrt{\frac{t}{T_1}}}$) which we do observe experimentally.

\subsection{Stretch exponential profile of the lifetimes}
\begin{figure}
\includegraphics[scale=0.7]{T1-stretch_exp}
\caption{Lifetime measurement for a case where all four classes are degenerate. Plain orange line correspond to a stretch-exponential fit and dashed red line to a simple exponential fit}
	\label{stretch}
\end{figure}
In the theory developed in \cite{choi_depolarization_2017}, the stretch exponential profile arises from the inhomogeneity of the distance from each NV centers to the closest fluctuators. If $\rho_{00}^s(t)$, the population in the $\ket{m_s=0}$ state for each NV centers evolving in the dark, follow a law of the form $\rho_{00}^s(t) \propto \exp(-\gamma t)$ where $\gamma$ is the individual depolarization rate of the spin; then, assuming an even spatial distribution of fluctuators, the authors of \cite{choi_depolarization_2017} show that the distribution in $\gamma$ should follow a law of the form $$\rho(\gamma) \approx \frac{e^{-\frac{1}{4 \gamma T_1}}}{\sqrt{4 \pi \gamma^3 T_1}} $$ where $\rho(\gamma)$ is the density of probability of $\gamma$.

Averaging then over all NV centers gives the stretch exponential profile observed from the ensemble :
$$ \rho_{00}^e(t) \propto \int_0^{+\infty} \rho(\gamma) e^{-\gamma t} d\gamma = e^{-\sqrt{\frac{t}{T_1}}} $$
where $ \rho_{00}^e(t)$ correspond to the average population in the $\ket{m_s=0}$ state for the ensemble of spins.

Fig. \ref{stretch} shows a lifetime measurement on a static micro diamond following the protocol described in Sec.III. Here all four classes are resonant with the applied microwave frequency, which corresponds to the maximum degree of degeneracy between the NV centers, and therefore the stronger modification of the lifetime induced by the resonant dipolar coupling. The signal we obtain was fitted using a stretch exponential profile  and a simple exponential profile. We can see that the stretch exponential profile ($R^2=0.981$) is in better agreement with the data than the exponential fit ($R^2=0.942$), in particular for the very short times (we expect the longer times to be dominated by the phonon-limited exponential lifetime).

Finally it should be noted that the stretch exponential profile arising from point-like depolarization sources is a relatively general result that has for example also been observed for the depolarization of NV centers induced by substitutional nitrogen (P1) defects in diamond \cite{hall_detection_2016}

\subsection{Scanning the degeneracy conditions}
\begin{figure}
\includegraphics[scale=0.3]{scan_CR}
\caption{Experimental results for a magnetic field scan with an electromagnet while a static magnetic field offset is applied with a permanent magnet (setup similar to the one presented in Fig.\ref{Optics}) \textbf{a)} Frequency Transitions for the $\ket{m_s=0} \to \ket{m_s=-1}$ transition of the four classes of NV centers as a function of the scanning magnetic field (data extrapolated from ODMR spectra) \textbf{b)} Change in photoluminescence of the NV$^-$ centers as a function of the scanning magnetic field. \textbf{c)} Example of 3 ODMR spectra for the 3 particular field values corresponding to the circled areas in subfigure a).}
	\label{scan_dege}
\end{figure}
The easiest way to probe the mechanism of dipolar-induced modification of the lifetime is to change the degeneracy conditions between the four classes of NV centers by tuning the magnetic field, as explained in Sec.I.
Because the NV spins can only exchange spin quanta when they are quasi-resonant, then by tuning the number of classes at degeneracy, you can modify the effective density of interacting NV centers, and therefore tune the depolarization effect.

An example of this is given in Fig.2 of the main text with the varying lifetime depending on the degeneracy condition, but another way to probe this effect is shown in Fig.\ref{scan_dege} : in this figure, we have observed the change in photoluminescence from a static microdiamond while changing the magnetic field in order to explore different degeneracy conditions. In order to do this, we need two sources of magnetic field : an electromagnet to scan the field and a permanent magnet to apply a magnetic field offset in an other direction (otherwise the magnetic field orientation with respect to the diamond axes would remain the same as the field is scanned).

In this particular case, we can see that as the magnetic field is scanned, it crosses three "degeneracy planes" (as described in Sec.I) : first a plane of the the $\{ 110 \}$ family at B=13 G, with a single degeneracy condition, then a plane of the $\{ 100 \}$ family at B=44 G where there is a simultaneous degeneracy condition for two pairs of classes, and then another plane of the the $\{ 110 \}$ family at B=76 G. We notice that each time a degeneracy between at least two classes of NV happens, we observe in Fig.\ref{scan_dege}b) a sharp decrease in photoluminescence. This is a signature of the change in lifetime of the ensemble of spins : indeed, the photoluminescence of NV ensembles is proportional to the average population in the $\ket{m_s=0}$ state, and the $\ket{m_s=0}$ population of the spins is the result of the competition between the polarization rate due to the green laser and the various depolarization mechanism. Increasing the depolarization rate of the spins will therefore decrease the overall luminosity.

@@

Another way to estimate the torque is to numerically solve the master equation of the system as depicted in Fig \ref{Torque1classe}. We find that under green excitation and in the presence of an external magnetic field, the spins will acquire a magnetization $\gamma_e \langle\hat{\mathbf S}\rangle$ which, under the low magnetic fields ($<$ 200 G) we are working at, will be oriented at an angle of 90$^\circ$ from the NV axis : $\langle \hat S_{z'} \rangle \approx 0$ and $\langle \hat S_\perp \rangle \neq 0$. This magnetization vanishes when the magnetic field is aligned with the NV center since there is no longer a transverse field responsible for the mixing of the eigenstates.

@@

The amplitude of the torque with respect to the magnetic field orientation at a B field amplitude of 100 G is represented in 1D in Fig \ref{Torque1classe}-c) where we can see a behavior very close to $|\sin(2\theta)|$, as found in Eq.\ref{torque_anal} through a perturbative approach. The same torque amplitude is represented in 2D in Fig \ref{Torque1classe} d). The two purple dots in the map correspond to the [111] direction when the magnetic field is aligned with the centers. The curvy line corresponds to the (111) plane. Importantly, the maximum torque value is $1\cdot 10^{-26}$ N.m for a single spin.

@@

Fig 3 a) in the main text, represents the same map as Fig \ref{Torque1classe}-c) but including the four NV centers, one in each of the possible [111] orientations. We can see that the maximum torque actually decreased to $3\cdot 10^{-28}$ N.m even though we increased the number of NV centers by four. This is due to the directional averaging of the torque generated by the four centers. The torque per NV center is decreased by more than two orders of magnitude when taking the directional averaging into consideration.
Fig 3 b) in the main text shows the same map, this time taking into account the modification of the spin lifetime due to cross-relaxations. The detailed model is presented in the section \ref{Simu}.
There are two things to note here : 
\begin{enumerate}
\item The maximum torque has increased by an order of magnitude compared to the previous case. It reached up to $3\cdot 10^{-27}$ N.m for four spins, so about $10^{-27}$ Nm per spin. Qualitatively, this is because cross-relaxation will lower the torque contribution of specific classes (the ones that get depolarized), meaning that the end result is closer to the single spin case (there is less directional averaging).
\item The change in magnetic torque is resonant, it happens only when different classes are brought to resonance. This can be seen by comparing Fig 3 b) of the main text to the $\{ 110 \}$ planes that were drawn in Fig. \ref{cristallo}. The change in the signal when scanning a magnetic field across a CR will be much sharper than the sinusoidal change in the spin-torque.
\end{enumerate}

@@

It should be noted that the main uncertainty comes here from the diamond size which can change the expected result by an order of magnitude.
